Bayesian Deep Gaussian Processes for Correlated Functional Data: \\ A Case Study in Cosmological Matter Power Spectra

### 2025-07-01 paper/main.tex
include Data Availability Statement
rm extra line break
add middle initial; rm extra Mpc

\title{Bayesian Deep Gaussian Processes for Correlated Functional Data: \\
        A Case Study in Cosmological Matter Power Spectra}
\author{Stephen A. Walsh\thanks{Corresponding author: Division of Natural Sciences, 
        Math and Technology, Elms College, {\tt walshst@elms.edu}} \and 
        Annie S. Booth\thanks{Department of Statistics, Virginia Tech} \and
        David Higdon\footnotemark[2] \and
        Jared Clark\footnotemark[2] \and
        Kelly Moran\thanks{Los Alamos National Laboratory} \and
        Katrin Heitmann\thanks{Argonne National Laboratory}}
\date{\today}

@@

For each cosmology, Mira-Titan contains a batch of 18 simulated power spectra: an inexpensive 
power spectrum estimated from perturbation theory ($y_p$), 16 power spectra estimated from 
low resolution simulations ($y_{\ell_r}, r \in \{1,\dots,16\}$), and one power spectrum from 
a high resolution simulation ($y_h$). 
These power spectra are represented as a function of wavenumber $k$, expressed in units of 
$\text{Mpc}^{-1}$ (inverse megaparsecs).  Following 
\cite{moran2023mira}, we model on the emulation space 
$\mathcal{P}(k)=\log_{10}\left(\frac{k^{1.5}P(k)}{2\pi^2}\right)$, where $P(k)$ represents the
original spectra. Output is available for $n=351$ values of $k$, spanning
$0.001 \leq k \leq 5 \text{ Mpc}^{-1}$, but each data type has unique wavenumber ranges
where it is deemed approximately unbiased. For $k<0.04 \text{ Mpc}^{-1}$, only $y_p$ provides a reliable 
estimate of the true spectra. $y_{\ell_r}$ are valid for $0.04 \leq k \leq 0.25 \text{ Mpc}^{-1}$, and 
$y_h$ is valid for $0.04 \leq k \leq 5 \text{ Mpc}^{-1}$.  Figure \ref{fig:plot_data} shows an example 
of the output for a particular cosmology. When we focus on $k$ values where $y_{\ell_r}$ 
is valid (center panel), the spatially correlated fluctuations of the low-resolution and 
high-resolution runs becomes apparent.

@@

Here we describe our first contribution: the use of a Bayesian hierarchical model 
to estimate the underlying spectrum and quantify uncertainty for a particular cosmology.
Let input $X$ represent the vector with elements $x_i = \log_{10}(k_i)$ for $i=1,\dots,n$.
In the previous section, we used lowercase $\bar{y}$ to denote 
the observed weighted average; now, we use $\bar{Y}$ to indicate the corresponding 
random variable within the statistical model.

$\bar{Y}$ is typically pre-processed
based on domain expertise.  Pre-processing for Mira-Titan was described in Section 
\ref{sec:data}; we will detail pre-processing 
for the CAMB suite in Section \ref{subsec:camb}.

@@

>>>>

\title{Bayesian Deep Gaussian Processes for Correlated Functional Data: \\
        A Case Study in Cosmological Matter Power Spectra}
\author{Stephen A. Walsh\thanks{Corresponding author: Division of Natural Sciences, 
        Math and Technology, Elms College, {\tt walshst@elms.edu}} \and 
        Annie S. Booth\thanks{Department of Statistics, Virginia Tech} \and
        David Higdon\footnotemark[2] \and
        Jared Clark\footnotemark[2] \and
        Kelly R. Moran\thanks{Los Alamos National Laboratory} \and
        Katrin Heitmann\thanks{Argonne National Laboratory}}
\date{\today}

@@

For each cosmology, Mira-Titan contains a batch of 18 simulated power spectra: an inexpensive 
power spectrum estimated from perturbation theory ($y_p$), 16 power spectra estimated from 
low resolution simulations ($y_{\ell_r}, r \in \{1,\dots,16\}$), and one power spectrum from 
a high resolution simulation ($y_h$). 
These power spectra are represented as a function of wavenumber $k$, expressed in units of 
$\text{Mpc}^{-1}$ (inverse megaparsecs).  Following 
\cite{moran2023mira}, we model on the emulation space 
$\mathcal{P}(k)=\log_{10}\left(\frac{k^{1.5}P(k)}{2\pi^2}\right)$, where $P(k)$ represents the
original spectra. Output is available for $n=351$ values of $k$, spanning
$0.001 \leq k \leq 5 \text{ Mpc}^{-1}$, but each data type has unique wavenumber ranges
where it is deemed approximately unbiased. For $k<0.04$, only $y_p$ provides a reliable 
estimate of the true spectra. $y_{\ell_r}$ are valid for $0.04 \leq k \leq 0.25$, and 
$y_h$ is valid for $0.04 \leq k \leq 5$.  Figure \ref{fig:plot_data} shows an example 
of the output for a particular cosmology. When we focus on $k$ values where $y_{\ell_r}$ 
is valid (center panel), the spatially correlated fluctuations of the low-resolution and 
high-resolution runs becomes apparent.

@@

Here we describe our first contribution: the use of a Bayesian hierarchical model 
to estimate the underlying spectrum and quantify uncertainty for a particular cosmology.
Let input $X$ represent the vector with elements $x_i = \log_{10}(k_i)$ for $i=1,\dots,n$.
In the previous section, we used lowercase $\bar{y}$ to denote 
the observed weighted average; now, we use $\bar{Y}$ to indicate the corresponding 
random variable within the statistical model.  
$\bar{Y}$ is typically pre-processed
based on domain expertise.  Pre-processing for Mira-Titan was described in Section 
\ref{sec:data}; we will detail pre-processing 
for the CAMB suite in Section \ref{subsec:camb}.

@@

\section{Data Availability Statement}
We provide data and reproducible code and an {\sf R} package for our 
Bayesian DGP.FCO model in a public git repository: \url{https://github.com/stevewalsh124/dgp.hm}
